Optimised Cost Considering Huffman Code for Biological Data Compression

### 2015-02-02 Article/Article.tex
discussion finished

\author{}

@@

\begin{abstract}
%% Text of abstract
Classical Huffman code has widely been used to compress biological datasets. Though a considerable reduction of size of data is obtained by classical Huffman code, yet, more efficient encoding is possible by treating binary bits differently considering requirement of transmission time, energy consumption, and likewise. A number of techniques already modified Huffman code algorithm to  obtain optimal prefix-codes for unequal letter costs in order to reduce overall transmission cost (time). 
%In the recent year with the wide use of cloud and the distributed application in our life, the data transmission become more challenging over the years. 
In this paper we propose a new approach to improve compression performance of one such extension, cost considering approach (CCA),   %the data transmission technique by improving the representation of Huffman code with the use of a 
by applying genetic algorithm for optimal allocation of the codewords to the symbols. 
%The work in this research is consists of two steps. 
The approach start with generating a set of codewords for different input symbols based on their frequencies by considering cost (length) of 0 and 1 as $\alpha$ and $\beta$,~$\alpha < \beta$. This step ensures optimal cost encoding, however, compression performance is not higher than classical Huffman code. 
%The aim of the codes generation method is to minimise the number of ones on the generated codes, which minimise the cost of the transmission. 
%To improve compression performance, the second part of the approach uses genetic algorithm to find an optimal allocation of generated codes to different symbols. 
The idea is to sacrifice some cost to minimise total number of bits, hence, the genetic algorithm works by giving penalty on cost. The performance of the approach is evaluated by applying it to compress some standard biological dataset and comparing the result with the results produced by classical Huffman code and standard cost considering Huffman code.
% The case study of the approach is the biological data, which present a good application area to data compression with the huge growth of the biological data bases. 
%The results are very promising and we have compared with the standard cost considering Huffman code and the original Huffman code.
\end{abstract}

\begin{keyword}
Data Transmission , Huffman code, Cost Considering, Genetic algorithm, Biological data.
%% keywords here, in the form: keyword \sep keyword

@@

\begin{algorithm}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.7]{Images/Drawing1.jpg}
\caption{Population Update for genetic algorithm}
\end{center}
\label{Fig2}
\end{figure}

\section{Results And Discussion}
\label{sec4}
The approach has been evaluated with different real genomic biological data, these genomes were downloaded from a recent version of The National Center for Biotechnology Information (NCBI) available on $(http://www.ncbi.nlm.nih.gov)$. We focused on the sequences alone, ignoring any header and any other exogenous information. In table~\ref{table3} the different data set are described with the size of each of them and the references on the biological data bank.
\begin{table}
\small
\label{table3}
\caption{Dataset description}
\begin{center}
\begin{tabular}{c  c  c c}
\toprule
$\textbf{Data sets}$ &$\textbf{Name}$ &	$\textbf{Size (MB)}$ &	$\textbf{Reference}$ \\\hline
Genome 1& Mycobacterium smegmatis &  6.66 & CP009496  \\\hline

@@

\multirow{2}{*}{Genome 16}& Danio rerio & \multirow{2}{*}{56.14} &\multirow{2}{*}{CM002885}\\ & chromosome 1 &\\
\hline
\multirow{2}{*}{Genome 17}&Homo sapiens & \multirow{2}{*}{76.64 } &\multirow{2}{*}{  CM000680  }\\    & chromosome 18 &\\
\hline
\multirow{2}{*}{Genome 18}&Homo sapiens & \multirow{2}{*}{74.92} &\multirow{2}{*}{CM000684   }\\ & chromosome 22&\\
\hline
\bottomrule
\end{tabular}
\end{center}
\end{table}

\begin{table}
\small
\label{table4}
\caption{GABM for comparison for cost and number of bits of different approaches with different datasets}
\begin{center}
\begin{tabular}{c | c c | c c | c c}
\toprule
$\textbf{Data sets}$ & \multicolumn{2}{c}{Huffman Algorithm} & \multicolumn{2}{c}{CCA }& \multicolumn{2}{c}{OCCA}\\\hline
 & $\textbf{Cost}$	& $\textbf{Bits}$ &	$\textbf{Cost}$&	$\textbf{Bits	}$&$\textbf{Cost}$&	$\textbf{Bits}$
\\\hline
Genome 1& 76787151  &	4.44	&67416213   & 4.92	& 67416213 & 4.82 \\\hline
Genome 2& 100425402 &  	5.81	& 88430665	& 6.51	& 88430665&  6.35 \\\hline
Genome 3& 75940155  &	4.39    &66745619	& 4.87&	66745619 & 4.79 \\\hline
Genome 4& 103552729 &	5.96    &90821835&	6.69 &	90821835  & 6.47\\\hline
Genome 5& 82234926  &	4.74	&71963876	& 5.27 &	71963876 &5.17 \\\hline
Genome 6& 83454842  &   4.81	&73038795	& 5.34	&73038795 & 5.24\\\hline
Genome 7& 92539488  &	5.36    &81416359   &5.93	&81416359 & 5.85\\\hline
Genome 8& 99613856  & 	5.74	&87102639   &6.41	&87102639&	6.30 \\\hline
Genome 9& 71876739  &	4.15    &62998800  &4.58 &	62998800 8&	5.50 \\\hline
Genome 10& 75324432 & 	4.36	&66084958  &4.81	&66084958&	4.73 \\\hline
Genome 11& 80766360 &	4.66    &70620666  &5.41&	70620666&	5.04 \\\hline
Genome 12& 74560825 &	4.31    &65359604  &4.75	&65359604&	4.67\\\hline
Genome 13& 71562941 & 	4.14	&62758225  &4.56	&	62758225 &	4.50\\\hline
Genome 14& 78261299 & 	4.51    &68354090  &4.8&	68354090& 4.89  \\\hline
Genome 15& 324439242& 	18.77    &286008420 & 20.66	 &	288866525 &	19.94 \\\hline
Genome 16& 703734840&  40.77    &618859291 & 45.08  &	625042576 &	43.18 \\\hline
Genome 17& 901032667&  52.08    &791840455 &	57.36 &	799757406 &	55.12 \\\hline
Genome 18& 983434816&  56.98    &867299889 &	62.42 &	875969422 & 60.70 \\\hline

\bottomrule
\end{tabular}
\end{center}
\end{table}

\begin{table}
\small
\label{table5}
\caption{Influence of penalty on bit minimisation}
\begin{center}
\begin{tabular}{c c c c c }
\toprule
$\textbf{Data sets}$ & $\textbf{Cost}$	& $\textbf{Bits}$ &		\textbf{$\lambda (\%)$} &		\textbf{Compression ratio $\%$} \\\hline
Genome 1 &	 70760174 &4.50& 4 & 32.72\\\hline
Genome 2 &	92010490& 5.91 & 3& 28.79\\\hline
Genome 3&	69421783& 4.47 &3& 26.84\\\hline
Genome 4&	96294638& 5.99 & 5& 29.85\\\hline
Genome 5&	74855668 &4.81& 5& 27.77\\\hline
Genome 6&	76738330& 4.86  &4& 28.10\\\hline
Genome 7&	84667949 &5.47& 3& 26.28\\\hline
Genome 8&	92416243 &5.76& 5& 30.35\\\hline
Genome 9&	66145783&4.21 & 4& 26.65\\\hline
Genome 10&	68751359& 4.44 &3& 26.63\\\hline
Genome 11&	72737300 &4.79& 2&26.08 \\\hline
Genome 12&	67981988 & 4.38&3& 26.63\\\hline
Genome 13&	65896779& 4.18 &4& 27.30\\\hline
Genome 14&	71762305 &4.55 &4&27.54 \\\hline
Genome 15&297188002 &19.31 & 3& 24.51\\\hline
Genome 16&643785655 & 41.68 & 3& 25.75 \\\hline
Genome 17&823506882 & 53.05 & 3& 30.78\\\hline
Genome 18&901515198 &58.99 & 3& 21.26\\\hline
\bottomrule
\end{tabular}
\end{center}
\end{table}
\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.7]{Images/Drawing4_1.jpg}
\caption{Convergence of Optimised cost considering Huffman algorithm}
\end{center}
\label{Fig5}
\end{figure}
The table~\ref{table4} presents the results founded by the basic Huffman code, cost considering algorithm and optimised cost considering algorithm. The result show that the number of bits of Huffman algorithm is the minimum number among the other algorithm but the cost is  very high. The cost considering algorithm improve the quality of the generated codes in terms of cost but the number of bits. The optimised cost considering algorithm try to find the best allocation of codes to frequencies while the cost constraint respected. The table 4 present the best founded number of bits with different penalty, for each genome we find the max number of useful penalty, after this value, increasing the penalty are in-useful (See figure 3), the number of bits achieve the minimum number but the cost stop decreasing (point 1 in figure 3) and this number of bits stabilized while the cost still increasing until it stabilized also (point 2 figure 3), after this point the cost and number of bits are stabilized.

>>>>

%\author{}
\author[a,b]{Youcef Gheraibia\corref{cor1}}
\ead{youcef.gheraibia@gmail.com}
\author[b]{Sohag Kabir\corref{cor2}}
\ead{s.kabir@hull.ac.uk}

@@

\address[a]{Department of Mathematics and Computer Science , University of Souk Ahras, Souk Ahras 41000, Algeria}

\address[b]{Department of Computer Science, University of Hull, Hull HU6 7RX, UK}

\cortext[cor1]{Corresponding author. Tel.: +0-000-000-0000 ; fax: +0-000-000-0000.}

\cortext[cor2]{Corresponding author. Tel.: +447405024667 ; }

\begin{abstract}
%% Text of abstract
Classical Huffman code has widely been used to compress biological datasets. Though a considerable reduction of size of data is obtained by classical Huffman code, yet, more efficient encoding is possible by treating binary bits differently considering requirement of transmission time, energy consumption, and likewise. A number of techniques already modified Huffman code algorithm to  obtain optimal prefix-codes for unequal letter costs in order to reduce overall transmission cost (time). 
%In the recent year with the wide use of cloud and the distributed application in our life, the data transmission become more challenging over the years. 
In this paper we propose a new approach to improve compression performance of one such extension, cost considering approach (CCA),   %the data transmission technique by improving the representation of Huffman code with the use of a 
by applying genetic algorithm for optimal allocation of the codewords to the symbols. 
%The work in this research is consists of two steps. 
The approach start with generating a set of codewords for different input symbols based on their frequencies by considering cost (length) of 0 and 1 as $\alpha$ and $\beta$,~$\alpha < \beta$. This step ensures optimal cost encoding, however, compression performance is not higher than classical Huffman code. 
%The aim of the codes generation method is to minimise the number of ones on the generated codes, which minimise the cost of the transmission. 
%To improve compression performance, the second part of the approach uses genetic algorithm to find an optimal allocation of generated codes to different symbols. 
The idea is to sacrifice some cost to minimise total number of bits, hence, the genetic algorithm works by giving penalty on cost. The performance of the approach is evaluated by applying it to compress some standard biological datasets and comparing the result with the results produced by classical Huffman code and cost considering Huffman code.
% The case study of the approach is the biological data, which present a good application area to data compression with the huge growth of the biological data bases. 
%The results are very promising and we have compared with the standard cost considering Huffman code and the original Huffman code.
\end{abstract}

\begin{keyword}
Data Compression, Huffman code, Cost Considering, Genetic algorithm, Biological data.
%% keywords here, in the form: keyword \sep keyword

@@

\begin{algorithm}[t]

@@

\begin{figure}[!thpb]
\begin{center}
\includegraphics[scale=0.7]{Images/Drawing1.jpg}
\caption{Population Update for genetic algorithm}
\end{center}
\label{Fig2}
\end{figure}

\section{Results And Discussion}
\label{sec4}
The effectiveness of the approach has been evaluated with different real genomic biological data, these genomes were downloaded from a recent version of The National Center for Biotechnology Information (NCBI) available on $(http://www.ncbi.nlm.nih.gov)$. We focused on the sequences alone, ignoring any header and any other exogenous information. In table~\ref{table3} the different data sets are described with the size of each of them and the references on the biological data bank.
\begin{table}[!thpb]
\small
\label{table3}
\caption{Dataset description}
%\begin{center}
\centering
\begin{tabular}{c  c  c c}
\toprule
$\textbf{Data sets}$ &$\textbf{Name}$ &	$\textbf{Size (MB)}$ &	$\textbf{Reference}$ \\\hline
Genome 1& Mycobacterium smegmatis &  6.66 & CP009496  \\\hline

@@

\multirow{2}{*}{Genome 16}& Danio rerio & \multirow{2}{*}{56.14} &\multirow{2}{*}{CM002885}\\ & chromosome 1 &\\
\hline
\multirow{2}{*}{Genome 17}&Homo sapiens & \multirow{2}{*}{76.64 } &\multirow{2}{*}{  CM000680  }\\    & chromosome 18 &\\
\hline
\multirow{2}{*}{Genome 18}&Homo sapiens & \multirow{2}{*}{74.92} &\multirow{2}{*}{CM000684   }\\ & chromosome 22&\\
\hline
\bottomrule
\end{tabular}
%\end{center}
\end{table}

\begin{table}[h]
\small
\label{table4}
\caption{GABM for comparison for cost and number of bits of different approaches with different datasets}
%\begin{center}
\centering
\begin{tabular}{c | c c | c c | c c}
\toprule
$\textbf{Data sets}$ & \multicolumn{2}{c}{Huffman Algorithm} & \multicolumn{2}{c}{CCA }& \multicolumn{2}{c}{OCCA}\\\hline
 & $\textbf{Cost}$	& $\textbf{Bits}$ &	$\textbf{Cost}$&	$\textbf{Bits	}$&$\textbf{Cost}$&	$\textbf{Bits}$
\\\hline
Genome 1& 76787151  &	4.44	&67416213   & 4.92	& 67416213 & 4.82 \\\hline
Genome 2& 100425402 &  	5.81	& 88430665	& 6.51	& 88430665&  6.35 \\\hline
Genome 3& 75940155  &	4.39    &66745619	& 4.87&	66745619 & 4.79 \\\hline
Genome 4& 103552729 &	5.96    &90821835&	6.69 &	90821835  & 6.47\\\hline
Genome 5& 82234926  &	4.74	&71963876	& 5.27 &	71963876 &5.17 \\\hline
Genome 6& 83454842  &   4.81	&73038795	& 5.34	&73038795 & 5.24\\\hline
Genome 7& 92539488  &	5.36    &81416359   &5.93	&81416359 & 5.85\\\hline
Genome 8& 99613856  & 	5.74	&87102639   &6.41	&87102639&	6.30 \\\hline
Genome 9& 71876739  &	4.15    &62998800  &4.58 &	62998800 8&	5.50 \\\hline
Genome 10& 75324432 & 	4.36	&66084958  &4.81	&66084958&	4.73 \\\hline
Genome 11& 80766360 &	4.66    &70620666  &5.41&	70620666&	5.04 \\\hline
Genome 12& 74560825 &	4.31    &65359604  &4.75	&65359604&	4.67\\\hline
Genome 13& 71562941 & 	4.14	&62758225  &4.56	&	62758225 &	4.50\\\hline
Genome 14& 78261299 & 	4.51    &68354090  &4.8&	68354090& 4.89  \\\hline
Genome 15& 324439242& 	18.77    &286008420 & 20.66	 &	288866525 &	19.94 \\\hline
Genome 16& 703734840&  40.77    &618859291 & 45.08  &	625042576 &	43.18 \\\hline
Genome 17& 901032667&  52.08    &791840455 &	57.36 &	799757406 &	55.12 \\\hline
Genome 18& 983434816&  56.98    &867299889 &	62.42 &	875969422 & 60.70 \\\hline

\bottomrule
\end{tabular}
%\end{center}
\end{table}

\begin{table}[t]
\small

\caption{Influence of penalty on bit minimisation}
%\begin{center}
\centering
\begin{tabular}{c c c c c }
\toprule
$\textbf{Data sets}$ & $\textbf{Cost}$	& $\textbf{Bits}$ &		\textbf{$\lambda (\%)$} &		\textbf{Compression ratio $\%$} \\\hline
Genome 1 &	 70760174 &4.50& 4 & 32.72\\\hline
Genome 2 &	92010490& 5.91 & 3& 28.79\\\hline
Genome 3&	69421783& 4.47 &3& 26.84\\\hline
Genome 4&	96294638& 5.99 & 5& 29.85\\\hline
Genome 5&	74855668 &4.81& 5& 27.77\\\hline
Genome 6&	76738330& 4.86  &4& 28.10\\\hline
Genome 7&	84667949 &5.47& 3& 26.28\\\hline
Genome 8&	92416243 &5.76& 5& 30.35\\\hline
Genome 9&	66145783&4.21 & 4& 26.65\\\hline
Genome 10&	68751359& 4.44 &3& 26.63\\\hline
Genome 11&	72737300 &4.79& 2&26.08 \\\hline
Genome 12&	67981988 & 4.38&3& 26.63\\\hline
Genome 13&	65896779& 4.18 &4& 27.30\\\hline
Genome 14&	71762305 &4.55 &4&27.54 \\\hline
Genome 15&297188002 &19.31 & 3& 24.51\\\hline
Genome 16&643785655 & 41.68 & 3& 25.75 \\\hline
Genome 17&823506882 & 53.05 & 3& 30.78\\\hline
Genome 18&901515198 &58.99 & 3& 21.26\\\hline
\bottomrule
\end{tabular}
%\end{center}
\label{table5}
\end{table}

\begin{figure}[hbt]
\begin{center}
\includegraphics[scale=0.7]{Images/Drawing4_1.jpg}
\caption{Convergence of Optimised cost considering Huffman algorithm}
\end{center}
\label{Fig5}
\end{figure}
Table~\ref{table4} presents the results obtained by the classical Huffman code, cost considering algorithm (CCA) and optimised cost considering algorithm (OCCA) without penalty on cost. The results show that the number of bits required by the classical Huffman algorithm to encode genomic data is the minimum among the other algorithms but the cost is  maximum. The cost considering algorithm improve the representation of the generated codes in terms of cost but the number of bits. However, still it compresses the data by X\% in the best case, Y\% in the worst case, and Z\% on an average. In terms of cost, in the best case the CCA improves cost over classical Huffman code by P\%, in the worst case by Q\%, and on an average by R\%.  The optimised cost considering algorithm tries to find the best allocation of codes to frequencies by giving penalty on cost. The outcome of this process is a fall in the total number of bits and a rise in the total cost. However, the cost is always lower than the cost incurred by the classical Huffman algorithm.  At first, the OCCA optimises number of bits without applying any penalty on cost. Afterwards, it continues giving penalty on cost ranging from 1\% to 10\% until a balance is found between total cost and bits. For each genome, a maximum amount of effective penalty is identified, after this maximum value, increasing the penalty is not helping any more to reduce number of bits, i.e., number of bits reach to minimum and cost to maximum.  The table \ref{table5} presents the best founded number of bits for different datasets with different effective penalty.  %are in-useful (See figure 3), the number of bits achieve the minimum number but the cost stop decreasing (point 1 in figure 3) and this number of bits stabilized while the cost still increasing until it stabilized also (point 2 figure 3), after this point the cost and number of bits are stabilized.
